Improved Approximation Algorithms for the Maximum Carpool Matching Problem

### 2017-01-30 src/local.tex
Merge Dror

The local search algorithm maintains a feasible matching throughout its execution
and operates in 
iterative manner where in each iteration it tries to find a better solution by
replacing a subset of the edges in the current solution with 
another (larger) subset of edges not in the solution.
The local search algorithm halts when no improvement can be done.
The algorithm is described in Algorithm~\ref{alg:local}.

\begin{algorithm}
\SetKw{True}{true}
\SetKw{False}{false}
\KwIn{$G = (V, A)$, $c : V \rightarrow \N$, $k$}
\KwOut{$M$}
$M \leftarrow \emptyset$								\\
\Repeat{done}{
\label{line:outerloop}
 	$done \leftarrow{}$ \True							\\
 	\For{$M' \in M : |M'| \leq k$}{
 		\For{$A' \in A \setminus M : |A'| = |M'| + 1$}{
			\If{$M \setminus M' \cup A'$ is feasible}{
				$M \leftarrow{} M \setminus M' \cup A'$	\\
				$done \leftarrow{}$ \False				\\
			}
		}
 	}
}
\Return{$M$}

\caption{
\label{alg:local}
Local Search}
\end{algorithm}

We claim that the local search algorithm halts in polynomial time.
Observed that in every iteration the algorithm either improves the value of the solution
by one or otherwise it terminates. 
Thus, after a maximum of $n$ iterations the algorithm stops.
In every iteration we examine all the subsets of edges of a fixed size and test for feasibility,
both these operations can be done in polynomial time - $O(n^{Ck + 1})$.

To analyze the approximation ratio achieved by the local search algorithm consider an 
arbitrary but fixed optimal matching $M^*$.
Given the solution outputted by the local search algorithm, 
$M$, 
we build the \emph{star graph} of the two solutions
in which each node represents a star from the optimal solution 
and edges exits between two nodes if
there is a star in $S(M)$ that intersect the two
corresponding stars of the optimal solution, 
formally $G^{M^*}_M = (S, E)$ where $\{v_i : S^*_i \in S(M^*) \}$ 
and 
$E = \{(v_i, v_j) : 
\exists S \in M,
S \cap S^*_i \neq \emptyset \land S \cap S^*_j \neq \emptyset,
\}$.
Figure~\ref{fig:conflict} depicts a star graph.

\begin{figure}[ht]
\centering
\subfloat[]{
\label{subfloat:graph}
\input{fig-sub-graph}
}
\subfloat[]{
\label{subfloat:conflict}
\input{fig-sub-conflict}
}
\caption[]{
\label{fig:conflict}
\subref{subfloat:input} 
An optimal matching depicted by the dashed red edges,
and another solution depicted by the solid green edges.  
\subref{subfloat:output}
The conflict graph, the upper left vertex corresponds 
to the star that contains the vertices a,b,c.
The numbers inside the vertices corresponds to $d_i$.   
}
\end{figure}

For a group of vertices, $U$, denote by $w^*(U)$ the sum of the degrees 
induced by the underlying graph of the optimal solution on this set, i.e.
$\sum_{u \in U} \dinout[M^*](u)$, 
and let $w(S)$ be the sum of the degrees 
induced by the underlying graph of the local search solution on this set, i.e.
$\sum_{u \in U} \dinout(u)$.

\begin{lemma}
\label{lemma:r}
If $k \geq \text{TODO}$ then in any connected induced graph of $G^{M^*}_M$, 
$G'= G^{M^*}_M[U]$, 
$
\frac{w(U)}{w^*(U)} 
\geq \frac{1}{2} + \frac{1}{2C} - \frac{1}{2C|U|}
$.
\end{lemma}

\begin{proof}
Consider the solution obtained from the $M$ by removing all the edges from $M$
that intersect the induced subgraph, and adding all the edges from $M^*$ that
intersect the induced subgraph.
It is easy to verify that this is a feasible solution.
We now bound from below the difference in the value of the solution.
The value decrease due to the removal of the edges by at most $w(U) - |U| + 1$,
and increase by exactly $\frac{w^*(U)}{2} \leq C |U|$.
We know that this difference can not be positive, or else, the local search
algorithm would not have been terminated.
Thus 
$\frac{w^*(U)}{2} \leq w(U) - |U| + 1$ 
and so
$
\frac{w(U)}{w^*(U)} 
\geq 
\frac{1}{2} + \frac{|U|}{w^*(U))} - \frac{1}{w^*(U))}  
\geq
\frac{1}{2} + \frac{1}{2C} - \frac{1}{2C|U|}
$
\end{proof}

@@

We now show that Theorem~\ref{theorem:local search} is tight.
Consider the example in Figure~\ref{fig:local search tight},
the example is for the special case when $k=11$ and $C=2$ but this example can
be generalized in a straightforward manner.
One can verify that as the graph in the example growth, 
the ratio between the optimal solution and the local search solution approaches
$\frac{3}{4}$.
\begin{figure}
\begin{center}
\caption{
\label{fig:local search tight}
A tight example for Theorem~\ref{theorem:local search}. 
The optimal solution is given by the red solid arcs while the local search
solution is given by the green, dashed arcs.
The solution can be improved by the local search algorithm only if it removes
all the edges. 
}
\input{fig-local-tight}
\end{center}
\end{figure}

>>>>

In this section we present a $(\frac{2c_{\max}}{c_{\max} + 1}
+ \eps)$-approximation algoithm for unweighted \carpool whose running
time is polynomial if $c_{\max} = O(1)$.

Let $k$ be an integer to be determined later.  The local search
algorithm maintains a feasible matching throughout its execution and
operates in iterative manner where in each iteration it tries to find
a better solution by replacing a subset of at most $k$ edges in the
current solution with another (larger) subset of edges not in the
solution.  The local search algorithm halts when no improvement can be
done.  The algorithm is described in Algorithm~\ref{alg:local}.

\begin{algorithm}
\SetKw{True}{true}
\SetKw{False}{false}
\KwIn{$G = (V, A)$, $c : V \rightarrow \N$, $k$}
\KwOut{$M$}
$M \leftarrow \emptyset$								\\
\Repeat{done}{
\label{line:outerloop}
 	$done \leftarrow{}$ \True							\\
 	\ForAll{$M' \subseteq M : |M'| \leq k$}{
 		\ForAll{$A' \subseteq A \setminus M : |A'| = |M'| + 1$}{
			\If{$M \setminus M' \cup A'$ is feasible}{
				$M \leftarrow{} M \setminus M' \cup A'$	\\
				$done \leftarrow{}$ \False				\\
			}
		}
 	}
}
\Return{$M$}

\caption{Local Search}
\label{alg:local}
\end{algorithm}

\todo[inline]{fix running time paragraph}

We claim that the local search algorithm halts in polynomial time.
Observed that in every iteration the algorithm either improves the
value of the solution by one or otherwise it terminates.  Thus, after
a maximum of $n$ iterations the algorithm stops.  In every iteration
we examine all the subsets of edges of a fixed size and test for
feasibility, both these operations can be done in polynomial time.

Observe that a vertex $v \in D_M \cup Z_M$ is the center of
a \emph{directed star} whose leaves are the passengers in the set
$P_M(v) = \set{u : (u,v) \in M}$.  ($P_M(v) = \emptyset$, for $v \in
Z_M$.)
%
Given a carpool matching $M$, we define $\calS(M)$ to be the set of
stars that are induced by $M$.  Denote by $V(S)$ the set of vertices
of a star, i.e., if $v$ is the center $S$, then $V(S) = \{v\} \cup
P_M(v)$.  Also, for $\calT \subseteq \calS(M)$, define
$V(\calT) \eqdf \bigcup_{S \in \calT} V(S)$.  Also, let $A(S)$ be the
edges of $S$, and define $A(\calT) \eqdf \bigcup_{S \in \calT} A(S)$.

To analyze the approximation ratio achieved by the local search
algorithm consider an arbitrary but fixed optimal matching $M^*$.
Given the solution outputted by the local search algorithm, $M$, we
build the \emph{star graph} of the two solutions in which each node
represents a star from the optimal solution, namely from $\calS(M^*)$,
and an edge exists between two nodes if there is a star in $\calS(M)$
that intersects the two corresponding stars of the optimal solution.
%
Formally $H = (\calS(M^*), E)$ where
\[
E = \set{(S^*_i, S^*_j) : \exists S \in \calS(M), 
         V(S) \cap V(S^*_i) \neq \emptyset \land
         V(S) \cap V(S^*_j) \neq \emptyset}
~.
\]
Figure~\ref{fig:conflict} depicts a star graph.

\begin{figure}[t]
\centering
\subfloat[]{
\label{subfloat:graph}
\input{fig-sub-graph}
}
\subfloat[]{
\label{subfloat:conflict}
\input{fig-sub-conflict}
}
\caption[]{
\subref{subfloat:input} 
An optimal matching depicted by the dashed red edges,
and another solution depicted by the solid green edges.  
\subref{subfloat:output}
The conflict graph, the upper left vertex corresponds 
to the star that contains the vertices a,b,c.
The numbers inside the vertices corresponds to $d_i$.   
}
\label{fig:conflict}
\end{figure}

\begin{lemma}
\label{lemma:r}
Let $\calT \subseteq \calS(M^*)$ that induces a connect subgraph of
$H$.  If $\abs{A(\calT)} \leq k$, then
\[
\frac{\deg_M(V(\calT))}{\deg_{M^*}(V(\calT))} 
\geq \half + \inv{2\cmax} - \inv{2\cmax\abs{\calT}}
~.
\]
\end{lemma}
\begin{proof}
Consider the solution $M'$ obtained from the $M$ by removing all the
edges from $M$ that intersect $V(\calT)$ and adding all the edges from
$M^*$ that intersect $V(\calT)$.  Observe that if an edge $(u,v)$ in
$M^*$ intersects $V(\calT)$, then $\set{u,v} \in V(\calT)$ by the
definition of the graph $H$.  Hence, $M'$ is feasible carpool
matching.

Since $\calT$ induces a connected subgraph of $H$, the removal of
edges in $M$ that intersect $V(\calT)$ decreased $\abs{M}$ by at most
$\deg_M(V(\calT)) - \abs{\calT} + 1$.
%
On the other hand, the increase in size is exactly
 $\half \deg_{M^*}(V(\calT)) \leq \cmax \abs{\calT}$.
%
Since $\abs{A(\calT)} \leq k$, we know that this difference can not be
positive, or else, the local search algorithm would not have been
terminated.  Thus 
\(
\half \deg_{M^*}(V(\calT)) \leq \deg_M(V(\calT)) - \abs{\calT} + 1
\),
and so
\[
\frac{\deg_M(V(\calT))}{\deg_{M^*}(V(\calT))}
\geq \half + \frac{\abs{\calT} - 1}{\deg_{M^*}(V(\calT))}
\geq \half + \frac{\abs{\calT} - 1}{2\cmax \abs{\calT}}
~,
\]
as required.
\end{proof}

\begin{lemma}
If $k \geq 24\cmax$, then $\abs{M} \geq (\frac{\cmax+1}{2\cmax}
- \frac{2}{k}) \cdot \abs{M^*}$.
\end{lemma}
\begin{proof}
Consider a maximal (with respect to set inclusion) connected component
of $H$ induced by the vertices in $\calT \subseteq \calS(M^*)$.
%
If $\abs{M \cap A(\calT)} \leq k$, then it must be that $\abs{M \cap
A(\calT)} = \abs{M^* \cap A(\calT)}$, since otherwise $M \cap
A(\calT)$ could be improved.

It remains to consider a maximal component $\calT$ such that If
$\abs{M \cap A(\calT)} > k$.  Observe that since the number of edges
in a $S \in \calS(M^*)$ is at most $\cmax$, it must be that
$\abs{V(\calT)} > \frac{k}{\cmax}$.
%
Given such a component $\calT$, we partition its vertex set into
several sets, each of size at least $\frac{k}{4\cmax}$.  We do this as
follows.  First construct any spanning tree of $\calT$ (say, by
running DFS).  It is well known that a tree of $t$ vertices can be
split into two subtrees each with at most $\floor{2t/3}+2$ vertices
(see, e.g., \cite{???}).  Moreover, if $t \geq 24$, each such subtree
contains at most $\frac{3}{4}t$.  Hence, we can decompose $\calT$
recursively into vertex sets whose sizes is between $\frac{k}{\cmax}$
and $\frac{k}{4\cmax}$, assuming that $\frac{k}{\cmax} \geq 24$.
%
Each such vertex set $\calX$ is connected and contains at most
$\frac{k}{\cmax}$ stars.  Hence $\abs{A(\calX)} \leq k$.  Due to
Lemma~\ref{lemma:r} and since $\abs{V(\calX)} \geq \frac{k}{4\cmax}$,
we have that
\[
\frac{\deg_M(V(\calX))}{\deg_{M^*}(V(\calX))} 
\geq \half + \inv{2\cmax} - \frac{2}{k}
~.
\]
Since $\deg_{M^*}(V(\calT)) = \sum_{\calX} \deg_{M^*}(V(\calX))$ and
$\deg_M(V(\calT)) = \sum_{\calX} \deg_M(V(\calX))$, it follows that
\[
\frac{\deg_M(V(\calT))}{\deg_{M^*}(V(\calT))}
\geq \half + \inv{2\cmax} - \frac{2}{k}
~,
\]
and thus $\abs{M \cap A(\calT)} \geq (\frac{\cmax + 1}{2\cmax}
- \frac{2}{k}) \abs{M^* \cap A(\calT)}$.
\end{proof}

\begin{corollary}
\label{cor:local}
There exists a $(\frac{2\cmax}{\cmax+1} +\eps)$-approximation
algorithm for \carpool, for every $\eps>0$.
\end{corollary}

\iffalse %%%%% Rmeoved

@@

\fi %%%%% End removed

\todo[inline]{Add a proper theorem}

We now show that Corollary~\ref{cor:local} is tight.  Consider the
example in Figure~\ref{fig:local search tight}, the example is for the
special case when $k=11$ and $C=2$ but this example can be generalized
in a straightforward manner.  One can verify that as the graph in the
example growth, the ratio between the optimal solution and the local
search solution approaches $\frac{3}{4}$.

\begin{figure}
\begin{center}
\input{fig-local-tight}
\caption{A tight example for Theorem~\ref{theorem:local search}. 
The optimal solution is given by the red solid arcs while the local
search solution is given by the green, dashed arcs.  The solution can
be improved by the local search algorithm only if it removes all the
edges.}
\label{fig:local search tight}
\end{center}
\end{figure}

### 2017-01-30 src/main.tex
Merge Dror

\documentclass[draft]{article}

\input{in/libs}
\input{in/def}
\input{in/tikz}

\input{in/title}
\input{in/authors}

\begin{document}
\maketitle

@@

\bibliographystyle{plain}
\bibliography{main}
% \resizebox{.9\textwidth}{!}{\input{tmp}}

>>>>

%\documentclass[11pt,draft]{article}
%\documentclass[a4paper,11pt]{article}
\documentclass[a4paper,UKenglish]{lipics-v2016}

\usepackage{microtype}%if unwanted, comment out or use option "draft"

\usepackage{amsmath,amssymb,mathtools,amsthm}
\usepackage[linesnumbered, ruled]{algorithm2e}
\usepackage{xspace}
\usepackage{subfig}
\usepackage{paralist}
%\usepackage{hyperref}
%\hypersetup{final}
%\usepackage[inline]{enumitem}

% DEBUG
%\usepackage{fullpage}
\usepackage{todonotes}
%\usepackage{lineno}
%\linenumbers

% EXTRA
%\usepackage{authblk}

\input{in/def}
\input{in/tikz}

\renewcommand{\floatpagefraction}{0.7}

\sloppy

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{Improved Approximation Algorithms for the
Maximum Carpool Matching Problem}

\titlerunning{Improved Approximation Algorithms for Maximum Carpool Matching}

\author[1]{Gilad Kutiel}
\author[2]{Dror Rawitz%
\thanks{Supported in part by the Israel Science Foundation (grant no.~497/14).}}

\affil[1]{Department of Computer Science, Technion, Haifa,
  Israel. \\ \texttt{gkutiel@cs.technion.ac.il}}
\affil[2]{Faculty of Engineering, Bar Ilan University, Ramat Gan,
  Israel. \\ \texttt{dror.rawitz@biu.ac.il}}

\authorrunning{G. Kutiel and D. Rawitz}

\Copyright{G. Kutiel and D. Rawitz}

\subjclass{F.2.2 Nonnumerical Algorithms and Problems}

\keywords{approximation algorithms, local search}

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Acces}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
% Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

@@

\subparagraph*{Acknowledgements.}

We thank David Adjiashvili and Reuven Bar-Yehuda for helpful
discussions.

%\bibliographystyle{plain}
\bibliographystyle{plainurl}% the recommended bibstyle
\bibliography{main}

### 2017-01-30 src/sub.tex
Merge Dror

In this section we show that the problem can be formulated as unconstrained
submodular maximization problem, and thus, it has a 2-approximation algorithm.

>>>>

In this section we show that the \carpool problem can be formulated as
unconstrained submodular maximization problem, and thus it has a
randomized $2$-approximation algorithm based on~\cite{BFNS15}.

Given a \carpool instance $(G = (V,A), c, w)$, consider a subset
$S \subseteq V$.  Let $M(S)$ be the maximum weight carpool matching
that satisfies $D_{M(S)} \subseteq S \subseteq V \setminus P_{M(S)}$,
namely $M(S)$ is the best carpool matching whose drivers belong to $S$
and whose passengers belong to $V \setminus S$.  In other words,
$M(S)$ is the maximum weight carpool matching that is contained in
$A \cap (V \setminus S) \times S$.
%
Given $S$, the carpool matching $M(S)$ can be computed in polynomial
time using maximum flow as shown in~\cite{??}.

Consider the function $\barw: 2^V \to \R$, where $\barw(S) \eqdf
w(M(S))$.  In the next lemma we prove that $\bar{w}$ is
a \emph{submodular set function}.  Recall that a function $f$ is
submdular if $f(S) + f(T) \geq f(S \cup T) + f(S \cap T)$ for every
two sets $S$ and $T$ in the domain of $f$.

\begin{lemma}
$\barw$ is submodular.
\end{lemma}
\begin{proof}
Consider any two subsets $S, T \subseteq V$.  We show that $\barw(S)
+ \barw(T) \geq \barw(S \cup T) + \barw(S \cap T)$.
%
Let $M(S \cup T)$ and $M(S \cap T)$ be the optimal carpool matchings
with respect to $S \cup T$ and $S \cap T$, respectively.
%
To prove the lemma we construct two feasible carpool matchings $M_S$
and $M_T$ such that $M_S \subseteq (V \setminus S) \times S$,
$M_T \subseteq (V \setminus T) \times T$, and $M_S \cup M_T = M(S \cup
T) \cup M(S \cap T)$.

First, add all the edges in $M(S \cup T)$ entering $S \setminus T$ to
$M_S$.  Similarly, add all the edges in $M(S \cup T)$ entering
$T \setminus S$ to $M_T$.  Observe that $\din[M_S](v) = \din[M(S \cup
T)](v) \leq c(v)$, for every $v \in S \setminus T$ and that
$\din[M_T](v) = \din[M(S \cup T)](v) \leq c(v)$, for every $v \in
T \setminus S$.

Next, add the edges in $M(S \cap T)$ leaving $T \setminus S$ to $S$
and add the edges in $M(S \cap T)$ leaving $S \setminus T$ to $T$.
%
It remains to distribute the edges leaving $V \setminus (S \cup T)$
and entering $S \cap T$ in both $M(S \cup T)$ and $M(S \cap T)$.  Note
that there may exist edges $(v,u)$, where $v \not\in S \cup T$, and
$u \in S \cap T$ such that $(v,u) \in M(S \cup T)$ and $M(S \cap T)$.
We refer to this edges as \emph{duplicate} edges.
%
We add all edges leaving $V \setminus (S \cup T)$ and entering $S \cap
T$ in $M(S \cap T)$ to $M_S$.  Notice that this is possible, since
after this addition we have that $\din[M_S](v) \leq \din[M(S \cap
T)](v) \leq c(v)$, for every vertex $v \in S \cap T$.
%
Then we add all duplicate edges in $M(S \cup T)$ to $M_T$.
%
The remaining edges are distributed between $M_S$ and $M_T$ without
violating capacities.  This can be done, since $\din[M(S \cup T)](v)
+ \din[M(S \cap T)](v) \leq 2c(v)$, for every $v \in S \cap T$.

The lemma follows, since $\barw(S) \geq w(M_S)$ and $\barw(T) \geq
w(M_T)$.
\end{proof}

Buchbinder et al.~\cite{BFNS15} presented a general randomized
$2$-approximation algorithm for unconstrained submodular maximization.

\begin{theorem}
There exists a randomized polynomial time $2$-approximation algorithm
for \carpool.
\end{theorem}
